---
title: "Data Exploration"
author: "Jose Rafael Dimayacyac"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
groundhog::groundhog.library("tidyverse", date = "2022-04-21")
knitr::opts_knit$set(root.dir = here::here())
```

## Introduction

In this document I will be exploring the measurement data for the 6961 bird species used in the later analysis. Specifically, I want to understand the dimensions of the data, determine which data points will be useful for later analyses, as well as confirm the conclusions from the paper from which the data is taken from.

## Loading

First, I will load the data and get an idea of what it looks like.

```{r load}
measurements <- readxl::read_xlsx("data/mass_predictions.xlsx")
glimpse(measurements)
```
The authors used multiple measurements to try to infer body mass, but it seems like they were mainly loaded in as characters instead of numerics, and that most values were not able to be measured. Now I need to decide which variables I will use for the clustering. 

## Data Cleaning

Now I will clean up the data; i.e., coerce character values to numerical wherever appropriate, and identify which variables I will ultimately need. 

```{r clean}
data <- measurements %>% 
  rename(Mass = `Weight (gram)`) %>%
  select(!c(`Source (Weight)`)) %>%
  distinct() %>%
  pivot_longer(cols = c(Mass:Carpometacarpus)) %>%
  mutate(value = as.numeric(value)) %>%
  pivot_wider(names_from = name, values_from = value) #Not working quite right yet

glimpse(data)
```

As shown by the output of the glimpse function, the types have now been corrected for. However, it should be noted that there are a lot of missing values for actual mass; which is probably why mass needed to be predicted. Now I want to pick a set of variables to use to cluster the data.

## Variable Selection

Here I select which variables I will use to plot the data

```{r select}
data %>%
  filter(Extinct == "yes") %>%
  select(`Scientific name`, Mass:Carpometacarpus) %>% 
  pivot_longer(cols = !c(`Scientific name`)) %>%
  group_by(name) %>%
  summarise(NotAvailable = sum(is.na(value)), Available = sum(!is.na(value))) 

data %>%
  filter(Extinct == "no") %>%
  select(`Scientific name`, Mass:Carpometacarpus) %>% 
  pivot_longer(cols = !c(`Scientific name`)) %>%
  group_by(name) %>%
  summarise(NotAvailable = sum(is.na(value)), Available = sum(!is.na(value)))
```

As shown by the table above, most species have a Mass value, but extinct species do not. This means I cannot use the Mass value for clustering. However, many species do have Tarsometatarsus and Body Length measurements. I will filter for data points that have values for both of these.  

## Filtering

Outside of the predicted mass of the birds, the authors were interested in another variable; the ability of the birds to fly. The original paper concluded that extinct birds tended to be larger, and that flightless birds also tended to be larger. I also wished to understand how much birds varied in mass between families. Here I analyze how birds vary in mass by the variables of extinction status, flightlessness, and phylogenetic taxa. 

```{r filter}
data %>%
  select(`Scientific name`, Order, Flightless, Extinct, Body_Length, Tarsometatarsus) %>%
  filter(!is.na(Body_Length), !is.na(Tarsometatarsus))

data %>%
  select(`Scientific name`, Order, Flightless, Extinct, Body_Length, Humerus) %>%
  filter(!is.na(Body_Length), !is.na(Humerus))

data_filtered <- data %>%
  select(`Scientific name`, Order, Flightless, Extinct, Humerus, Tarsometatarsus) %>%
  filter(!is.na(Humerus), !is.na(Tarsometatarsus))
```

As shown by the above tibbles, species with body length measurements do not overlap with species with Tarsometatarsus measurements. This is the same for body length and humerus measurements. Therefore, I will chose to use the measurements of Humerus and Tarsometatarsus. Additionally, I am interested in Extinction status, flightlessness, and Order.

## Data cleaning

Finally, I produce a cleaned and filtered file to perform clustering on.
```{r clean}
write_csv(data_filtered, file = "data/data_clean.csv")
```